---
title: 'R Primer I: Refresher'
output:
  html_document: default
  html_notebook: default
---

```{r setup, cache = FALSE, echo = FALSE}
# Don't stop compiling the notebook when an error occurs
knitr::opts_chunk$set(error = TRUE)
```

This notebook provides a short refresher on R basics to quickly get started working in R. 

# Vectors and Matrices 

Let's first use the R prompt interactively for simple calculations:
```{r}
1 + 1
```

## Creating Vectors

Variables are assigned with the `<-` operator. While it is also possible to use `=` as the assignment operator, this is
avoided in standard R coding practice to help distinguish variable assignments from function parameters.
```{r}
x <- 1:10
x
```
The `:` operator is a simple way to create sequences in R, the data structure thus created is a vector. 
When arithmetic computations are performed on vectors, they are applied element-wise. For example, a multiplication of a
vector by a scalar results in all elements of the vector being multiplied.
```{r}
x * 2
```

To harness vectorized computation, it is useful to know the basic ways to create sequences:
```{r}
y <- rep(c(1, 2), 5)
y
x * y

y <- rep(c(1, 2), each = 5)
y
x * y

seq(0, 10, by = 2)
seq(0, 10, length = 5)
seq_along(y)
```

## Accessing elements of vectors

Elements within vectors can be selected in multiple ways, with either one of:

* positive numeric indices select the elements at the given position, with indexing starting at 1:
```{r}
x[1]
x[2:4]
x[c(2, 5, 3)]
```

* negative numeric indices select all elements except the ones a the given positions:
```{r}
x[-(2:4)]
```

* a boolean index vector of the same length as the original vector:
```{r}
x[rep(c(TRUE, FALSE), 5)]
```

* a character vector containing names of the elements:
```{r}
names(x) <- letters[seq_along(x)]
x
x[c('d', 'b', 'c')]
```

Names that do not exist produce `NA`'s
```{r}
x[c('a', 'foo', 'bar', 'b')]
```

It is not possible to mix multiple types of indexing, an error will occur if one tries to do so:
```{r}
x[-2:4]
```

## Matrices

A Matrix is a two-dimensional structure, with all elements being of the same type (e.g. numeric):
```{r}
x <- matrix(1:12, nrow = 3)
x
```

Elements from a matrix can be selected in a similar manner to vectors, with row and column indices separated by a comma.
All of the indexing methods for vectors also work for matrices.
```{r}
x[2, 3]
x[3, -2]
```

To select entire rows and columns, leave the column or row index empty.
```{r}
x[1, ]
x[, 2:3]
```

# Data Frames and Tibbles

## Creating and using data frames
More complex data can be stored in `data.frame`s and their more modern version, tibbles. The main difference between
a matrix and a data.frame or a tibble is that the latter two can have columns of different types.
```{r}
chol_df <- read.csv('cholesterol.csv')
```

To see the contents of a `data.frame`, we could just type its name as we have done with the other variables so far. 
However, this will print the whole dataset, which is more information than needed most of the time and particularly 
inconvenient for large datasets...
```{r}
chol_df
```

Instead, one can get a simple overview of the data using
```{r}
str(chol_df)
```

## Accessing elements of a data frame

As for vectors and matrices, there are multiple ways of accessing elements of a data.frame
```{r}
chol_df[1:10, ]
chol_df[, 2:3]
chol_df[, 'Time']
chol_df[, c('Time', 'Cholesterol')]
```

The `$` operator provides a shorthand access to entire columns:
```{r}
chol_df$Time
chol_df[['Time']]
```
To select rows based on their values in one of the columns, an ad-hoc boolean vector can be calculated, for example
using the comparison `==`. Like all other comparisons, this will applied elementwise and return a vector.
```{r}
chol_df[chol_df$Time == 1, ]
```

## Writing data frames to files

We can then also write data in much the same way as it was read from the disk:
```{r}
write.csv(chol_df, file = 'temp.csv', row.names = FALSE)
```

## Using tibbles

As noted above, the "tibble" is a more modern alternative to the `data.table`. It behaves very similarly, but with
slight improvements. The R package `tibble` can be loaded to use tibbles...
```{r}
library(tibble)
```
...but there is also a more convenient way to do so by loading the `tidyverse` 'Metapackage' which loads a collection of
packages, including `tibble`.
```{r}
library(tidyverse)
```
The `read_csv` function is the `tidyverse` alternative to `read.csv`, and it returns a tibble instead of a 
`data.frame`. Note the `_` instead of the `.` in the function name: this is the naming convention in the tidyverse. In 
contrast to `read.csv`, character columns are not automatically converted to factors, and a report of the column types
is printed after reading a file:
```{r}
chol <- read_csv('cholesterol.csv')
```

By default, only the first 10 rows and only as many columns as fit on the screen are printed. In addition, the dataset
dimensions and the column types are displayed as additional information.
```{r}
chol
```

We can manually change the number of rows printed by calling the `print` function explicitely and using the parameter
`n`.
```{r}
print(chol, n = 40)
```

If we need to print the whole dataset, `n` can also be set to infinity:
```{r}
print(chol, n = Inf)
```

Note that a tibble (class `tbl`) is also at the same time a `data.frame`.
```{r}
class(chol)
```

Other ways of getting summary information of the dataset work as with `data.frame`s
```{r}
dim(chol)
str(chol)
```

## Accessing elements of a tibble

In general, a tibble can be used much in the same way as a `data.frame`
```{r}
chol[1:10, ]
chol[, 2:3]
```

```{r}
chol[, 'Time']
```

Note that when using `[` to create a subset, this always returns another tibble, in contrast to `data.frame`s, which 
will automatically convert the result into a vector if one of the dimensions is 1:
```{r}
chol_df[, 'Time']  # data.frame
```

To return a column as a vector, we have to do so explicitely with either
```{r}
chol[['Time']]  # tibble
```
or 
```{r}
chol$Time
```

## Reading and writing files in the tidyverse

As with `read_csv`, there is also a tidyverse function for writing datasets. In addition to being twice as fast, it
never writes rownames, so we can omit the `row.names` parameter that we used in `write.csv`.
```{r}
write_csv(chol, path = 'temp.csv')
```

The tidyverse read/write functions, including the `read_csv` and `write_csv` functions encountered above, can be found
in the package `readr`, which supports all kinds of text file inputs. 
```{r}
help(package = readr)
```

Other specific file types from other programs can be imported using the tidyverse packages `readxl` (for Microsoft Excel
files) and `haven` (for SPSS, SAS, and Stata).
```{r}
help(package = readxl)
help(package = haven)
```

R can also work directly with a database interface, see the package `DBI`, even using `dplyr` functions for database 
queries if the package `dbplyr` is installed (`dplyr` will be covered in Part II of this primer).
