---
title: 'R Primer II: Data Wrangling'
output:
  html_document: default
  html_notebook: default
---

```{r setup, cache = FALSE, echo = FALSE}
# Don't stop compiling the notebook when an error occurs
knitr::opts_chunk$set(error = TRUE)
```


# dplyr and the tidyverse

On of the most efficient ways to manipulate data in R is with the package `dplyr`, which is included in `tidyverse`. 
`dplyr` functions are not restricted to `tibble`s, they also work on `data.frame`s as well as other data structures.
```{r}
library(tidyverse)
chol <- read_csv('cholesterol.csv')
```

One can apply `dplyr` functions one by one...
```{r}
chol_filtered <- filter(chol, Time == 1)
chol_filtered
chol_selection <- select(chol_filtered, ID, Cholesterol)
chol_selection
chol_ordered <- arrange(chol_selection, Cholesterol)
chol_ordered
```

...or all at once with a rather convoluted nested function call, which is shorter but also more difficult to read: 

```{r}
arrange(select(filter(chol, Time == 1), ID, Cholesterol), Cholesterol)
```

One of the main advantages of `dplyr` comes into play with the `%>%` ("pipe") operator. Here, the functions used are
listed more intuitively in the order in which they are called, and all parameters immediately follow the function name. 
The output of the previous step is always used as the first parameter (unless specified otherwise, see below).
```{r}
chol %>%
  filter(Time == 1) %>%
  select(ID, Cholesterol) %>%
  arrange(Cholesterol)
```

Another reason to use `dplyr` is performance. Operations on `tibble`s are already faster than their equivalent on
`data.frame`s...

```{r}
big_df <- data.frame(x = rnorm(1e8), y = rnorm(1e8))
big_tibble <- tibble(x = rnorm(1e8), y = rnorm(1e8))
system.time(temp <- big_df[big_df$x < 0, ])
system.time(temp <- big_tibble[big_tibble$x < 0, ])
```

...but using the `dplyr` function is still more efficient (on both `tibble`s and `data.frame`s):
```{r}
system.time(temp <- big_df %>% filter(x < 0))
system.time(temp <- big_tibble %>% filter(x < 0))
```

Since we won't use these rather large datasets anymore, we can free up the memory they use by deleting them.
```{r}
rm(big_df, big_tibble, temp)
```

As a side note: while garbage collection can usually be relied upon to be done automatically, it is sometimes useful to
know that the function `gc` can be used to manually trigger the garbage collector, especially when handling large data.
```{r}
gc() # Explicit garbage collection
```

## Overview of dplyr functions

`dplyr` has functions for filtering rows...
```{r}
chol %>% filter(Cholesterol < 5)
```

...and selecting columns in a variety of ways.
```{r}
chol %>% select(Cholesterol)
chol %>% select(1:3)
chol %>% select(-Cholesterol)
chol %>% select(Margarine:Cholesterol)
chol %>% select(starts_with('M'))
chol %>% select(contains('gar'), ends_with('ol'))
```

`select` and `rename` can both define new variables, but the difference lies in which columns are returned
```{r}
chol %>% select(Ch = Cholesterol)
chol %>% rename(Ch = Cholesterol)
```

To add new colums to a dataset, use `mutate`:
```{r}
chol %>% mutate(Cholesterol_2 = Cholesterol - 5)
```

Sorting is done with `arrange`:
```{r}
chol %>% arrange(ID)
```

```{r}
chol %>% arrange(-Cholesterol)
```

Note that `-Cholesterol` is just an elementwise multiplication of the column `Cholesterol` by `-1`. Within the arguments
of dplyr functions, column names can be used like variables.
```{r}
chol %>% arrange(ID, -Cholesterol)
```

Aggregating over multiple rows is done with the function `summarize`:
```{r}
chol %>% summarize(mean_cholesterol = mean(Cholesterol))
```

Summaries such as this can also be computed using groups on the dataset:
```{r}
chol %>%
  group_by(Time) %>%
  summarize(mean_cholesterol = mean(Cholesterol))

chol %>%
  group_by(Time, Margarine) %>%
  summarize(mean_cholesterol = mean(Cholesterol)) %>%
  arrange(Margarine, Time)
```

We have now seen the basic functions that are used to manipulate data using `dplyr`, as well as the basic pipe operator 
`%>%` to make the code more readable.

## magrittr and pipes

There are a couple more useful pipe operators in the package `magrittr`
```{r}
library(magrittr)
```

It is very common to write results back to the original data:

```{r}
head(chol)
chol <- chol %>% arrange(desc(Cholesterol))
head(chol)
```

The `%<>%` pipe can be used as a shorthand for this. 
```{r}
head(chol)
chol %<>% arrange(ID, Time)  # shorthand for chol <- chol %>% arrange(ID, Time)
head(chol)
```

Usually, the pipe operator passes the output from the left hand side (LHS) as the first argument to the function on the
right hand side (RHS). If it is to be used in a different position, that position can be specified with a `.`:
```{r}
chol %>% lm(Cholesterol ~ Margarine * Time, data = .)
```

The `%$%` pipe operator can be used to access the columns of the output of the LHS by their names, as if they were
variables, in the expression on the RHS.
```{r}
chol %$% lm(Cholesterol ~ Margarine * Time)
```

Finally, the tee-pipe `%T>%` pipes its results into the next two functions, the output of the next function being 
ignored by the second function:
```{r}
chol %>%
  filter(ID == 1) %T>%
  (function(x) cat(paste('Using', nrow(x), 'Rows\n')))  %>%  # The output of this function is ignored
  summarize(x = mean(Cholesterol))  # This uses the output of two steps before
```

## Interlude: anonymous functions

Let's have a more detailed look at that function definition. Usually, when defining a function, that function gets named
by assigning the function to a variable:
```{r}
print_nrow <- function(x) {
  cat(paste('Using', nrow(x), 'Rows\n'))
}

chol %>% print_nrow()
```

We can, however, also use anonymous functions by defining them as we go:
```{r}
chol %>% (function(x) cat(paste('Using', nrow(x), 'Rows\n')))
```

Note that for use with the `%>%` operator (and for that use only), the function definition must be in parenthesis. 

Other typical examples of using anonymous functions in R is in apply calls (for faster loops)

In this example, we define a function that counts the number of unique values and apply it to each column of `chol`. 
`lapply` applies a function to each element of a list (or column of a `data.frame`/`tibble` or similar structure) and 
returns a list (`lapply` - "l" for "list").
```{r}
lapply(chol, function(x) {
  res <- length(unique(x))
  return(res)
})
```

`sapply` is similar to `lapply`, but tries to simplify the structure of the return value to a matrix or a vector, 
depending on the dimension of the output of the function (`sapply` - "s" for "simplify").
```{r}
sapply(chol, function(x) {
  res <- length(unique(x))
  return(res)
})
```

# Data wrangling in practice

Let's use `dplyr` to a practical case of preparing a dataset. First, two of the columns are recoded into factors. The 
mutation can directly be applied to the dataset, thus we can use the `%<>%` pipe.
```{r}
chol %<>% mutate(Time = as.factor(Time),
                 Margarine = as.factor(Margarine))
```

Assume we want to train a model where cholesterol values from multiple time points need to be considered simultaneously.
One way of doing this would be to create three sub-datasets containing the Cholesterol values of one time point each and
then joining these datasets:
```{r}
chol1 <- chol %>% filter(Time == 1) %>% select(-Time) %>% rename(T1 = Cholesterol)
chol2 <- chol %>% filter(Time == 2) %>% select(-Time) %>% rename(T2 = Cholesterol)
chol3 <- chol %>% filter(Time == 3) %>% select(-Time) %>% rename(T3 = Cholesterol)
chol_joined <- chol1 %>% 
                 left_join(chol2, by = c('ID', 'Margarine')) %>% 
                 left_join(chol3, by = c('ID', 'Margarine'))
chol_joined
```

This is, however, a rather common task, so there is a more efficient way of doing it. The function `spread` from `tidyr`
(another `tidyverse` package) performs this exact restructuring of a dataset. In this case, we want to retain the
original data as well as the modified dataset, so we assign the result to a new variable:
```{r}
chol_spread <- chol %>% spread(Time, Cholesterol)
colnames(chol_spread)[3:5] <- paste0('T', 1:3)
chol_spread
```

The inverse of the `spread` operation is called `gather`:
```{r}
chol_spread %>% gather('Time', 'Cholesterol', T1:T3)
```

Another rather typical task is creating a shifted version of a vector:
```{r}
lag(1:4)
lag(1:4, 1)
lag(1:4, 2)
```

Such a lag function could for example be used to compute the changes in cholesterol over time:
```{r}
chol %<>% group_by(ID) %>% 
  arrange(Time) %>% 
  mutate(Cholesterol_Diff = Cholesterol - lag(Cholesterol)) %>%
  ungroup() %>%
  arrange(ID, Time)

chol
```


# Nonstandard evaluation in tidyverse functions

The way that tidyverse functions accept object names as parameters -- without them being defined in the environment
where they are called, to be instead adaptively interpreted by the function during its evaluation -- is called
"Nonstandard evaluation" (NSE). While it makes for nicer usage in the interactive shell (among other advantages), it can
cause problems when programming with these functions:

```{r}
variables <- c('Margarine', 'Time')
lapply(variables, function(variable) {
  cholesterol_aggregated <- (chol %>% 
                              group_by(variable) %>%
                              summarize(median_cholesterol = median(Cholesterol)))[['median_cholesterol']]
})
```

This does not work because `group_by` expects its argument (`variable`) to be a column name. What we would like to do is
hand it a variable containing the column name instead. For these cases, each of these functions has a "sibling" with an
underscore at the end for bypassing NSE and instead using standard evaluation (SE).
```{r}
variable <- 'Time'

chol %>% select(variable)
chol %>% select('Time')
chol %>% select_('Time')
chol %>% select_(variable)
```

Using this SE-variant of the function, we can now fix the loop we tried to implement above:

```{r}
variables <- c('Margarine', 'Time')
lapply(variables, function(variable) {
  cholesterol_aggregated <- (chol %>% 
                              group_by_(variable) %>%
                              summarize(median_cholesterol = median(Cholesterol)))[['median_cholesterol']]
})
```
