---
title: "R Primer III - Exploratory Data Analysis"
output:
  html_document: default
  html_notebook: default
---

This part of the primer focuses on graphics for exploratory data analysis. Its goal is to show off four main types of
graphics and how they can be used to explore a dataset.

# Setting up the dataset

The graphics functions we use will come from the package `ggplot2`, another package of the `tidyverse`:
```{r}
library(tidyverse)
library(magrittr)
```

We will use data from the 2017 Stack Overflow developer survey, but to keep things simple we will only use a small
number of the variables of the survey[^1].

[^1]: The example presented here is based on an analysis published in the Stack Overflow blog: 
https://stackoverflow.blog/2017/06/15/developers-use-spaces-make-money-use-tabs/
```{r}
so_full <- read_csv('survey_results_public.csv')
dim(so_full)
so %>% select(Respondent, Country, DeveloperType, YearsCodedJob, TabsSpaces, Salary)
str(so)
```

## Handling categorical variables
In addition, we will load the `forcats` package for easier handling of categorical data.
```{r}
library(forcats)
```

The package `forcats` is used to rearrange and recode the factor variables for the purposes of this demonstration. 
The `forcats` functions can be recognized in the code below by their prefix: 
* `fct_` for functions that change the values in the factor
* `lvls_` for functions that only change the labels and ordering of the factor levels
```{r}
countries <- c('United States', 'India', 'United Kingdom', 'Germany', 'Canada')
so %<>% mutate(Country = fct_other(Country, keep = countries)) %>%
  mutate(Country = lvls_reorder(Country, c(5, 3, 4, 2, 1, 6)))

so %<>% filter(!is.na(TabsSpaces)) %>% mutate(TabsSpaces = lvls_reorder(as.factor(TabsSpaces), c(2, 3, 1)))

so %<>% 
  mutate(YearsCodedJob = fct_collapse(as.factor(YearsCodedJob), 
                                      '< 5y'  = c('Less than a year', 
                                                  '1 to 2 years', 
                                                  '2 to 3 years', 
                                                  '3 to 4 years', 
                                                  '4 to 5 years'),
                                      '6-10'  = c('5 to 6 years',
                                                  '6 to 7 years',
                                                  '7 to 8 years',
                                                  '8 to 9 years',
                                                  '9 to 10 years'),
                                      '11-15' = c('10 to 11 years',
                                                  '11 to 12 years',
                                                  '12 to 13 years',
                                                  '13 to 14 years',
                                                  '14 to 15 years'),
                                      '15+'   = c('15 to 16 years', 
                                                  '16 to 17 years',
                                                  '17 to 18 years',
                                                  '18 to 19 years',
                                                  '19 to 20 years',
                                                  '20 or more years'))) %>%
  mutate(YearsCodedJob = lvls_reorder(YearsCodedJob, c(3, 4, 1, 2)))
```

A simple yet effective plot to visualize numerical variables is the boxplot -- each variable is represented by a box 
stretching from the 25%-quantile to the 75%-quantile, with a line marking the median (or 50%-quantile). In this example,
we are interested in the question whether there is a relationship between a programmer's preferences for tabs or spaces 
and his salary.
```{r, warning=FALSE}
g <- ggplot(so, aes(x = TabsSpaces, y = Salary, fill = TabsSpaces)) + geom_boxplot(notch = TRUE, na.rm = TRUE)
plot(g)
```

The function call is pretty simple and helps establish the "grammar" used by the `ggplot2` package (actually,
"grammar of graphics" is what the "gg" in "ggplot" stands for). The first argument of `ggplot` is the dataset to be 
used, and its second argument is the aesthetics parameter -- this has to be an object produced by the function
`aes`, and it defines which of the variables of the dataset are used in which role. In this case, we want `TabsSpaces` 
to be represented on the x axis, `Salary` on the y axis and the fill color of the boxes determined by `TabsSpaces`. 

The aesthetics define the basic setup of the plot, which is more or less independent of the type of visualization. The
latter is defined by `geom_boxplot`, which, as the name suggests, draws boxplots, and is "added" to the plot with the
`+` operand. Using this operand, we can modify existing plots, as we will see in the next example. The plot definition
is then saved in the variable `g`, and drawn using the `plot` function[^2].

[^2]: We could also use the function `print` instead of `plot` to the same effect. `print` is also implicitly called 
whenever the result of a calculation is not stored to a variable, meaning that just typing `g` in the console will also
result in the plot being drawn (via the implicit `print` call).

The last thing to note on the boxplots is the notches drawn around the median using the parameter `notch = TRUE`, which 
indicate confidence intervals for the median. In this graphic, we can see that the confidence intervals of the medians 
of all three groups are much smaller than the difference between the medians of the "Spaces" box on the one hand and the
"Tabs" and "Both" boxes on the other, indicating that this difference is much larger than the sampling uncertainty of
the survey. We can thus conclude that the difference is very unlikely to be due to change alone. Developers using spaces
instead of tabs earn significantly more than those who prefer tabs.

In our next example, we will draw a barplot:
```{r}
g <- so %>% 
  group_by(TabsSpaces) %>% 
  summarize(Salary = median(Salary, na.rm = TRUE)) %>% 
  ggplot(aes(x = TabsSpaces, y = Salary, fill = TabsSpaces)) + 
  geom_col(na.rm = TRUE)
print(g)
```

Note that the call of the function `ggplot` looks identical to the one for the boxplots, the different plot type is
determined by the `geom_col` function that creates a barplot. In addition, since the barplot shows only one value per
level of TabsSpaces (in contrast to the boxplot that shows the distribution), we need to aggregate the values
beforehand. We use the median as a robust measure. What we see in this plot is basically a different visualization of
the three medians we already saw in the boxplots.

Things get more interesting if we split the results up by country to see whether the effect is seen across countries in
the same way, or whether this effect occurs only in some countries. To this end, we calculate the aggregates (medians)
by country and split the plot into subplots (facets) along the variable `Country` using the function `facet_wrap`. The
`scales` parameter says that the plots need not have the same scale on the y axis.
```{r}
g <- so %>% 
  group_by(Country, TabsSpaces) %>% 
  summarize(Salary = median(Salary, na.rm = TRUE)) %>% 
  ungroup() %>%
  ggplot(aes(x = TabsSpaces, y = Salary, fill = TabsSpaces)) + 
  geom_col(na.rm = TRUE) +
  facet_wrap(~Country, scales = 'free')
print(g)
```

The difference between the Tabs and Spaces groups is robust across countries, though the difference seems to be slightly
less pronounced in Germany and most pronounced in India. The main lesson here though is that geography (i.e. regional
preferences for tabs or spaces) is not the reason for the salary differences. Let's see whether programming experience
could be behind it.

To see the differences across different levels of programming experience, we will use a line chart. The usage looks very
similar to the multiple barplots, except that we now use `YearsCodedJob` instead of `TabsSpaces` as the x axis variable
and use `TabsSpaces` as the grouping variable. The grouping is not done in subplots as in the case of the barplots, but
within one plot, using the `group` aesthetic.
```{r}
g <- so %>% 
  filter(!is.na(YearsCodedJob)) %>%
  group_by(TabsSpaces, YearsCodedJob) %>% 
  summarize(Salary = median(Salary, na.rm = TRUE)) %>% 
  ungroup() %>%
  ggplot(aes(x = YearsCodedJob, y = Salary, group = TabsSpaces, col = TabsSpaces)) + 
  geom_line()
print(g)
```

We can thus note that the salary gap between space and tab users is more or less consistent across experience levels. 

Maybe different developer types with different salaries and perhaps different tab/space preferences can explain the gap
then? We can explore this using a point chart, which also introduces two more `ggplot` functions: `geom_abline` to draw
a line in the plot, and `geom_text` to annotate points. Finally, we will introduce the scaling functions for the x and y
axes, which in this graphic are continuous, thus we use `scale_x_continuous` and `scale_y_continuous`:

```{r}
library(ggrepel)  # provides the function geom_text_repel for prettier label placement
g <- so %>% 
  filter(TabsSpaces %in% c('Tabs', 'Spaces')) %>%
  mutate(DeveloperType = sapply(strsplit(DeveloperType, ';'), function(x) x[[1]])) %>%
  filter(!is.na(DeveloperType)) %>%
  group_by(DeveloperType, TabsSpaces) %>%
  filter(n() >= 50) %>%   # eliminate groups with few responses
  summarize(Salary = median(Salary, na.rm = TRUE)) %>%
  spread(TabsSpaces, Salary) %>%
  ggplot(aes(x = Tabs, y = Spaces)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, col = 'red') + 
  geom_text_repel(aes(label = DeveloperType)) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(labels = scales::comma)
print(g)
```

The red line gives us a reference for parity between tab and space users, and all points being above the line means that
among all developer types, the salaries of space users are consistently higher than those of tab users[^3].

[^3]: Our results in this plot differ slightly from the analysis as published on the Stack Overflow blog -- this is due 
to our simplified handling of the multi-input field `DeveloperType`, using only the first input of that field as our
classification and thus eliminating the overlap between the users of each type (which would lead to the analysis
pipeline less straightforward to read).

We have thus seen that developers that use spaces earn consistently more than developers using tabs, which leads us to 
our closing remark for this part of the primer: when performing statistical analyses, never confuse correlation with 
causation. Developers are very likely not paid more simply *because* they are using spaces instead of tabs, nor are they
likely to prefer spaces over tabs *because* they earn more. Rather, it seems more likely that there is a third factor 
correlated to both a preference for spaces over tabs *and* a higher salary that might explain the difference, but the 
jury is still out which factor that might be. 